Owl Occupancy in Three of El Salvador's Protected Areas from 2003 through 2013

### 2021-02-28 manuscript_working/ms_working_RMD.Rmd
Last RMD version, switching to word

Nancuchiname Forest (797 ha) is a Protected Natural Area located 95 km southeast of San Salvador in the coastal alluvial plain along the Lempa River (Fig. 1; Table 1).  Its elevation ranged from sea level to 12 m a.s.l.  The area contained deciduous and semi-deciduous forest, second growth vegetation, native species reforestation and a dike surrounded by secondary growth.  In Nancuchiname Forest, we established one survey route (N-1) in alluvial forest along the dike and the other route (N-2) in a semi-deciduous alluvial forest.

@@

The broadcast calls varied between routes, but were consistent at each route and station across years. Specifically, we broadcast the vocalizations of Pacific Screech-Owls (*Megascops cooperi*), Mottled Owls (*Ciccaba virgata*), Crested Owls (*Lophostrix cristata*), Black-and-white Owls (*C. nigrolineata*), Spectacled Owls (*Pulsatrix perspicillata*), Pacific Screech-Owls, Mottled Owls, Crested Owls, Black-and-white Owls, and Spectacled Owls at stations 1 through 10, respectively, in all routes in El Imposible NP (EI-1 and EI-2) and Nancuchiname Forest (N-1 and N-2). We broadcast the vocalizations of Whiskered Owls (*M. trichopsis*), Mottled Owls, Fulvous Owls (*Strix fulvescens*), Stygian Owls (*Asio stygius*), Great Horned Owls (*Bubo virginianus*), Whiskered Owls, Mottled Owls, Fulvous Owls, Stygian Owls, and Great Horned Owls at stations 1 through 10, respectively, in both routes of Montecristo NP (M-1 and M-2).

@@

Like many owl species, neotropical owls are difficult to accurately census due to their cryptic nature and nocturnal habits. Owls vocalize to communicate with the same species, to threaten or provoke other species, and to delimit territory [@Johnsgard:1988]. Distinguishing individuals by non-invasive means, such as vocal traits, is preferable when species are rare, sensitive to handling, elusive, or when other techniques are unfeasible [@Terry:2005]. Thus, the vocalizations of crepuscular and cryptic animals such as owls are commonly leveraged in field surveys because they can be diagnostic and are often considered more practically attainable than visual identification [@Konig:1999].

Vocalization surveys for nocturnal birds of prey fall generally into two categories. Spontaneous or “passive” surveys involve listening for animal calls with no prior stimulus, whereas provocation or “broadcast” surveys are conducted by listening after a period of playing a recording of animal vocalizations [e.g., @Baumgardt:2019].  Our surveys coupled both approaches, with spontaneous surveys preceeding broadcast surveys. Overall, the probability of detection was fairly low for all species (median values $\le 0.50$, even though we used locally-recorded owl vocalizations for our broadcast surveys when possible. In Guatemala, @Gerhardt:1989 found that a series of Mottled Owl vocalizations obtained from a non-local ornithological laboratory could not generate a response from Guatemalan owls. We found that the probability of detecting most owls increased after we broadcast the vocalizations of Mottled Owls, Pacific Screech-Owls, Crested Owls, and Spectacled Owls. Conversely, owls were less likely to respond after we played Great Horned Owl calls. Owls are sensitive to interactions with other owls [@Enriquez:1997; @Baumgardt:2019] and may change their response to broadcast calls depending on their life stage (e.g., nesting versus courtship) [@Flesch:2007], so we recommend that surveyors carefully consider the intra- and inter-species interactions that may affect detectability after playing specific bird vocalizations.

Globally, an estimated 75% of the nearly 250 species of owls are associated with dense and undisturbed forests [@Mikkola:2012].  @Perez-Leon:2017 stated that within El Salvador, natural ecosystems such as cloud, deciduous, riparian, and pine-oak forests had the most diverse owl populations, and our findings supported the importance of maintaining protected areas of El Salvador. We found that over the 10-year period of our study, owl species richness and single-species occupancy for three owls were all generally stable in the protected natural areas we investigated. Our study demonstrates the importance of maintaining natural forested areas and diverse habitat heterogeneity in a country under immense climate and population pressures.

>>>>

Nancuchiname Forest (797 ha) was a Protected Natural Area located 95 km southeast of San Salvador in the coastal alluvial plain along the Lempa River (Fig. 1; Table 1).  Its elevation ranged from sea level to 12 m a.s.l.  The area contained deciduous and semi-deciduous forest, second growth vegetation, native species reforestation and a dike surrounded by secondary growth.  In Nancuchiname Forest, we established one survey route (N-1) in alluvial forest along the dike and the other route (N-2) in a semi-deciduous alluvial forest.

@@

The broadcast calls varied between routes, but were consistent at each route and station across years. Specifically, we broadcast the vocalizations of Pacific Screech-Owls (*Megascops cooperi*), Mottled Owls (*Ciccaba virgata*), Crested Owls (*Lophostrix cristata*), Black-and-white Owls (*C. nigrolineata*), Spectacled Owls (*Pulsatrix perspicillata*), Pacific Screech-Owls, Mottled Owls, Crested Owls, Black-and-white Owls, and Spectacled Owls at stations 1 through 10, respectively, in all routes in El Imposible NP (EI-1 and EI-2) and Nancuchiname Forest (N-1 and N-2). We broadcast the vocalizations of Whiskered Owls (*M. trichopsis*), Mottled Owls, Fulvous Owls (*Strix fulvescens*), Stygian Owls (*Asio stygius*), Great Horned Owls (*Bubo virginianus*), Whiskered Owls, Mottled Owls, Fulvous Owls, Stygian Owls, and Great Horned Owls at stations 1 through 10, respectively, in both routes of Montecristo NP (M-1 and M-2). We used locally-recorded owl vocalizations for our broadcast surveys when possible, because @Gerhardt:1989 found that a series of Mottled Owl vocalizations obtained from a non-local ornithological laboratory could not generate a response from Guatemalan owls.

@@

Like many owl species, neotropical owls are difficult to accurately census due to their cryptic nature and nocturnal habits. Owls vocalize to communicate with the same species, to threaten or provoke other species, and to delimit territory [@Johnsgard:1988]. Distinguishing individuals by non-invasive means, such as vocal traits, is preferable when species are rare, sensitive to handling, elusive, or when other techniques are unfeasible [@Terry:2005]. Our surveys involved periods of passive listening followed by a broadcast call designed to trigger owl vocalization. Overall, the probability of detection was fairly low for all species (median values $\le 0.50$) and was generally lower during the passive listening periods pre-broadcast. Owls are sensitive to interactions with other owls [@Enriquez:1997; @Baumgardt:2019] and may change their response to broadcast calls depending on their life stage (e.g., nesting versus courtship) [@Flesch:2007], so we recommend that surveyors carefully consider the intra- and inter-species interactions that may affect detectability after playing specific bird vocalizations.

Globally, an estimated 75% of the nearly 250 species of owls are associated with dense and undisturbed forests [@Mikkola:2012].  @Perez-Leon:2017 stated that within El Salvador, natural ecosystems such as cloud, deciduous, riparian, and pine-oak forests had the most diverse owl populations, and our findings supported the importance of maintaining protected areas of El Salvador. We found that over the 10-year period of our study, overall owl species richness and single-species occupancy for three owls were generally stable in the protected natural areas we investigated. We also found that the probability of detecting most owls increased after we broadcast the vocalizations of Mottled Owls, Pacific Screech-Owls, Crested Owls, and Spectacled Owls. Conversely, owls were less likely to respond after we played Great Horned Owl calls. Our study demonstrates the importance of maintaining natural forested areas and diverse habitat heterogeneity in a country under considerable climate and population pressures.
